### 2018-03-10 paper/background.tex
Finish S2

\definition[Minimality]{A build system is \emph{minimal} if it executes tasks at
most once per build and only if they transitively depend on files that changed
since the previous build.}
\vspace{2mm}

@@

\noindent
The new version uses the \cmd{INDIRECT} function, which allows us to reference a
cell indirectly by a text string that is not necessarily known in advance. The
current implementation of \Excel recomputes indirect references in every
build~\cite{excel_recalc}. This clearly violates the minimality property: if
\cmd{C1=1} and the user modifies \cmd{A2}, \Excel will recompute \cmd{B1},
potentially triggering further unnecessary recalculation even though \cmd{B1}
does not transitively depend on \cmd{A2}.

@@

Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
systems track changes of inputs and intermediate results, executing dependent
tasks whenever they change, but \Excel can also track changes in the tasks
themselves: if a formula is modified, \Excel will recompute it and propagate
the changes. Self-tracking is uncommon in software build systems, where one
often needs to manually initiate a full rebuild even if just a single task has
changed. We further discuss self-tracking in~\S\ref{sec-tracking-aspects}.

@@

Note that it is not always possible to know the dependencies of a task upfront,
or \emph{statically}. To demonstrate this, let us add another task to the above
makefile:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{makefile}
release.tar: main.exe README
    tar -cf release.tar main.exe README
\end{minted}
\vspace{1mm}

\noindent
So far so good: if either \cmd{main.exe} or \cmd{README} is dirty, \Make will
recreate the release archive. But now imagine that we need to specify the files
that go into \cmd{release.tar} in a text file \cmd{release.txt}. The dependencies
of \cmd{release.tar} are no longer known statically: they \emph{depend on the
content} of \cmd{release.txt}, which might not even exist before the build
starts, e.g. if it is produced by concatenating input files \cmd{release-bins.txt}
and \cmd{release-docs.txt}. Alas, makefiles cannot express such \emph{dynamic
dependencies}, requiring ad hoc workarounds such as \emph{build phases}, which
are problematic~\cite{hadrian}.

\Shake was developed specifically to address the issue of dynamic
dependencies~\cite{mitchell2012shake} and it excels at handling them without
sacrificing the minimality requirement. Here is how one can express the task
for producing \cmd{release.tar} from the example discussed
in~\S\ref{sec-background-make}.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
"release.tar" %> \_ -> do
    need ["release.txt"]
    files <- lines <$> readFile "release.txt"
    need files
    system "tar" $ ["-cf", "result.tar"] ++ files
\end{minted}
\vspace{1mm}

\noindent
We first declare the static dependency on \cmd{release.txt}, then read its
content (a list of files) and depend on it, dynamically. Finally, we specify the
command to produce the resulting archive. Crucially, the archive will only be
rebuilt if one of the dependencies (static or dynamic) has changed.

\begin{figure}[h]
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
\vspace{-1mm}
\caption{Dependency graph produced after the previous build.}
\vspace{3mm}
\end{subfigure}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
\vspace{-1mm}
\caption{Since \cmd{docs.txt} is dirty, we rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering a new dependency.}
\end{subfigure}
\caption{Dynamic dependencies example: create \cmd{README} and add it to the
list of release documents \cmd{docs.txt}.\label{fig-shake}}
\end{figure}

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency graph produced in the previous build to detect
which files are dirty -- a powerful idea with a long history of use in
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting}~\cite{acar2007selfadjusting} \emph{computations} (see a
discussion in related work~\S\ref{sec-related}). Second, instead of abandoning
and deferring the computation of tasks whose rediscovered dependencies are dirty
(as \Excel does), \Shake \emph{pauses} their computation until the dependencies
are brought up-to-date. We refer to this build algorithm as \emph{recursive}.

@@

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The `Metadata' column refers to the \emph{auxiliary
build information} persistently stored between builds by all build systems:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that. In principle a single dirty bit is sufficient, as will be
    demonstrated in~\S\ref{sec-implementations}.
    \item \Excel stores the dirty bit as well as the calculation chain.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores all dependency graphs discovered in previous builds
    annotated with file hashes, and the content-addressable cache.
\end{itemize}

@@

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||l|c|c|c||l|l}
\hline
$\!\!$Build system & Dependencies & Minimal & Cutoff & Cloud & Metadata                       & Algorithm   $\!\!$\\\hline
$\!\!$\Make        & Static       & Yes     & No     & No    & File modification times        & Topological $\!\!$\\
$\!\!$\Excel       & Dynamic      & No      & No     & No    & Dirty cells, calculation chain & Chain, defer$\!\!$\\
$\!\!$\Shake       & Dynamic      & Yes     & Yes    & No    & Last dependency graph          & Recursive   $\!\!$\\
$\!\!$\Bazel       & Static       & Yes     & Yes    & Yes   & All dependency graphs, cache   & Topological $\!\!$\\\hline
\hline
\end{tabular}
\vspace{2mm}
\caption{Summary of build system differences.\label{tab-summary}}
\vspace{-7mm}
\end{table}

\subsection{Common vocabulary for build systems}
\label{sec-background-vocabulary}

In this section we introduce a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets. We will use this vocabulary throughout the rest of
the paper.

\emph{Keys} are used to distinguish \emph{values}. In software build systems
keys are typically filenames, e.g. \cmd{main.c}, whereas values are file
contents (a C program source code in this case). In spreadsheets keys are cell
names, e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In Haskell code, we will use type variables \hs{k}
and \hs{v} to denote keys and values, respectively.

We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for efficient
tracking and sharing of build results, where \hs{Hash} is an abstract data type
equipped with the equality test.

A \emph{store} associates keys to values. It is convenient to assume that a store
is total, i.e. it contains a value for every possible key. We therefore also
assume that the type of values is capable of encoding values corresponding to
non-existent keys (missing files or empty cells). This, in particular, suggests
that it is possible to \emph{depend on the absence of a value} which is a useful
feature for build systems. In addition to usual \hs{get} and \hs{put} operations,
some stores support the \hs{getHash} operation implemented more efficiently than
by hashing the result of \hs{get}. For example, GVFS (Git Virtual File
System)~\cite{gvfs} stores file hashes and downloads file contents only on demand.

Some values must be provided by the user as \emph{input}. For example,
\cmd{main.c} can be edited by the user who relies on the build system to
compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
but are produced in the process of turning inputs into outputs.

>>>>

\definition[Minimality]{A build system is \emph{minimal} if it executes tasks at
most once per build and only if they transitively depend on files that changed
since the previous build.}\label{def-minimal}
\vspace{2mm}

@@

\noindent
The new version uses the \cmd{INDIRECT} function, which allows us to reference a
cell indirectly by a text string that is not necessarily known in advance. The
current implementation of \Excel recomputes indirect references in every
build~\cite{excel_recalc}. This clearly violates the minimality
property~\ref{def-minimal}: if \cmd{C1=1} and the user modifies \cmd{A2}, \Excel
will recompute \cmd{B1}, potentially triggering further unnecessary
recalculation even though \cmd{B1} does not transitively depend on \cmd{A2}.

@@

Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
systems only track changes of inputs and intermediate results, but \Excel can
also track changes in the tasks themselves: if a formula is modified, \Excel
will recompute it and propagate the changes. Self-tracking is uncommon in
software build systems, where one often needs to manually initiate a full
rebuild even if just a single task has changed. We further discuss
self-tracking in~\S\ref{sec-tracking-aspects}.

\begin{figure}[h]
\vspace{-1mm}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
\caption{Dependency graph produced after the previous build.}
\end{subfigure}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
\caption{Since \cmd{docs.txt} has changed, we rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering a new dependency.}
\end{subfigure}
\vspace{-1mm}
\caption{Dynamic dependencies example: create \cmd{README} and add it to the
list of release documents \cmd{docs.txt}.\label{fig-shake}}
\vspace{-2mm}
\end{figure}

@@

\Shake was developed to solve the issue of dynamic
dependencies~\cite{mitchell2012shake} and it excels at them without sacrificing
the minimality requirement. To demonstrate this, we elaborate the \Make example
from~\S\ref{sec-background-make} by adding the following files whose
dependencies are shown in Fig.~\ref{fig-shake}(a):

\begin{itemize}
    \item \cmd{LICENSE} is an input text file containing the project license.
    \item \cmd{release.txt} is a text file listing all release files. This file
    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt}
    that list all binary and documentation files of the project.
    \item \cmd{release.tar} is the release archive built by executing the
    command \cmd{tar} on the release files.
\end{itemize}

The dependencies of \cmd{release.tar} are not known statically: they are
determined by the content of \cmd{release.txt}, which might not even exist
before the build. Makefiles cannot express such dependencies, requiring ad hoc
workarounds such as \emph{build phases}, which are problematic~\cite{hadrian}.
Here is how one can express the task for producing \cmd{release.tar} using \Shake.

\begin{minted}[xleftmargin=10pt]{haskell}
"release.tar" %> \_ -> do
    need ["release.txt"]
    files <- lines <$> readFile "release.txt"
    need files
    system "tar" $ ["-cf", "result.tar"] ++ files
\end{minted}

\noindent
We first declare the static dependency on \cmd{release.txt}, then read its
content (a list of files) and depend on it, dynamically. Finally, we specify the
command to produce the resulting archive. Crucially, the archive will only be
rebuilt if one of the dependencies (static or dynamic) has changed. For example,
if we create another documentation file \cmd{README} and add it to
\cmd{docs.txt}, \Shake will appropriately rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering the new dependency, as illustrated in
Fig.~\ref{fig-shake}(b).

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency from the previous build to detect which files need
to be rebuilt. This is a powerful idea with a long history of
use in \emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting}~\cite{acar2007selfadjusting} \emph{computations} (see a
discussion in related work~\S\ref{sec-related}). Second, instead of abandoning
and deferring the execution of tasks whose newly discovered dependencies have
not yet been built (as \Excel does), \Shake \emph{pauses} their execution until
the dependencies are brought up-to-date. We refer to this build algorithm as
\emph{recursive}.

@@

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The `Metadata' column refers to the \emph{auxiliary
build information} persistently stored between builds by all build systems:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that. In principle a single \emph{dirty bit} per file is
    sufficient, as will be demonstrated in~\S\ref{sec-implementations}.
    \item \Excel stores a dirty bit per cell as well as the calculation chain.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores all dependency graphs discovered in previous builds
    annotated with file hashes, and the content-addressable cache.
\end{itemize}

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||l|l||l|c|c|c}
\hline
$\!$Build system$\!$& Metadata                       & Algorithm    & Dependencies & Minimal & Cutoff & Cloud$\!$\\\hline
$\!$\Make       $\!$& File modification times        & Topological  & Static       & Yes     & No     & No   $\!$\\
$\!$\Excel      $\!$& Dirty cells, calculation chain & Reordering   & Dynamic      & No      & No     & No   $\!$\\
$\!$\Shake      $\!$& Last dependency graph          & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
$\!$\Bazel      $\!$& All dependency graphs, cache   & Topological  & Static       & Yes     & Yes    & Yes  $\!$\\\hline
\hline
\end{tabular}
\vspace{2mm}
\caption{Summary of build system differences.\label{tab-summary}}
\vspace{-7mm}
\end{table}

@@

### 2018-03-10 paper/compute.tex
Finish S2

\begin{minted}[xleftmargin=10pt]{haskell}
type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
\end{minted}

@@

>>>>

\begin{minted}[xleftmargin=10pt]{haskell}
type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
\end{minted}

@@

\subsection{Common vocabulary for build systems}
\label{sec-background-vocabulary}

In this section we introduce a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets. We will use this vocabulary throughout the rest of
the paper.

\emph{Keys} are used to distinguish \emph{values}. In software build systems
keys are typically filenames, e.g. \cmd{main.c}, whereas values are file
contents (a C program source code in this case). In spreadsheets keys are cell
names, e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In Haskell code, we will use type variables \hs{k}
and \hs{v} to denote keys and values, respectively.

We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for efficient
tracking and sharing of build results, where \hs{Hash} is an abstract data type
equipped with the equality test.

A \emph{store} associates keys to values. It is convenient to assume that a store
is total, i.e. it contains a value for every possible key. We therefore also
assume that the type of values is capable of encoding values corresponding to
non-existent keys (missing files or empty cells). This, in particular, suggests
that it is possible to \emph{depend on the absence of a value} which is a useful
feature for build systems. In addition to usual \hs{get} and \hs{put} operations,
some stores support the \hs{getHash} operation implemented more efficiently than
by hashing the result of \hs{get}. For example, GVFS (Git Virtual File
System)~\cite{gvfs} stores file hashes and downloads file contents only on demand.

Some values must be provided by the user as \emph{input}. For example,
\cmd{main.c} can be edited by the user who relies on the build system to
compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
but are produced in the process of turning inputs into outputs.
